Very simplified likelihood

### 2024-03-01 inst/ejs-sample.tex
start to write up for ejs

\begin{abstract}

Here a simplification of the Plackett-Luce likelihood functions for
order statistics is proposed.  We consider a field of $n+1$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Placket-Luce strength of $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  This device
furnishes a reasonable, intuitive, and computationally tractable
likelihood function that I apply to a range of applications including
Formula 1 motor racing and the International Mathematical Olympiad.
The results are difficult to obtain with other methods.
\end{abstract}

@@

\begin{equation}
\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n} 
+ K + \mathcal{O}\left(n^{-2}\right)
\end{equation}

Using $\mathcal{S}=\log(B-1)-B\frac{r+1}{n}$ as an approximate support
thus incurs an error of only $\mathcal{O}(n^{-2})$.  Observing that
$\partial^2\mathcal{S'}/\partial B^2=-(B-1)^{-2}<0$, the evaluate
would be unique; we have
$\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
alternatively $\hat{a}=\frac{n}{n+r+1}$.  I present some numerical
results below.

@@

\begin{figure}[t]
\includegraphics[width=4in]{ninelikes}  % produced by very_simplified_likelihood.Rmd
\caption{Likelihood functions for $r=0(1)8$\label{ninelikes}}
\end{figure}

@@

Formula 1 motor racing is an important and prestigious motor sport
\citep{codling2017,jenkins2010}.  Season ranking is based on a points
allocation system wherein competitors are awarded points based on race
finishing order; points accumulate additively.  The overall
competition winner is the competitor who accumulates the most points
after the final race.  However, as argued by Hankin~\cite{hankin2023},
the drivers' {\em ranks} are statistically sufficient for analysis of

The methods discussed above may be used to assess Bradley-Terry
strengths of Formula 1 racing drivers.  Let us consider 2018 season as
an example and first consider Lewis Hamilton.

Lewis Hamilton is a

Pierre Gasly and Esteban Ocon have a long-standing rivalry and
comparisons between them typically compare disparate metrics such as
number of podiums, and highest grid position in a season.

\begin{acks}[Acknowledgments]
The authors would like to thank the anonymous referees, an Associate
Editor and the Editor for their constructive comments that improved the
quality of this paper.
\end{acks}

>>>>

\begin{abstract}
Here a simplification of the Plackett-Luce likelihood functions for
order statistics is proposed.  We consider a field of $n+1$ entities
that are ordered in some way [the preferred example is competitors in
  a running race, ordered by time of crossing the finishing line].
One entity is of particular interest to us: the ``focal entity'', or
focal competitor.  We assign Placket-Luce strength of $a$ to the focal
competitor and $b$ to all others; we require $a+b=1$.  The associated
inference problem appears to have a wide range of applications and I
present some analysis of datasets drawn from Formula 1 motor racing
and community initiative Parkrun.
\end{abstract}

@@

\begin{equation}\label{asymptotic}
\log\mathcal{L}_{r,n}(a)=\log(B-1)-B\frac{r+1}{n}
+ K + \mathcal{O}\left(n^{-2}\right)
\end{equation}

Using $\mathcal{S}=\log(B-1)-B\frac{r+1}{n}$ as an approximate support
thus incurs an error of only $\mathcal{O}(n^{-2})$.  Observing that
$\partial^2\mathcal{S'}/\partial B^2=-(B-1)^{-2}<0$, the evaluate
would be unique; we have
$\hat{B}=\frac{n+r+1}{r+1}+\mathcal{O}\left(n^{-1}\right)$, or
alternatively $\hat{a}=\frac{n}{n+r+1}$.  For repeated observations
$r_i,n_i$, $1\leqslant i\leqslant m$ we would have
\begin{equation}
  \hat{a} = 1-
  \frac{\frac{r_1+1}{n_1}+\cdots+\frac{r_m+1}{n_m}}{m}
  = 1-\overline{\left(\frac{r+1}{n}\right)}.
\end{equation}

(that is, one minus the mean ratio of rank to number of [cloned]
competitors).  I present some numerical results below.

@@

\begin{figure}[t]
\includegraphics[width=4in]{ninelikes}  % ninelikes.pdf produced by very_simplified_likelihood.Rmd
\caption{Likelihood functions for $r=0(1)8$\label{ninelikes}}
\end{figure}

@@

Here I present three examples that showcase likelihood
function~\ref{likeforrn3}.  Formula 1 education, parkrun

\subsection{Park run}

``Parkrun'' is a distributed community initiative that organises
weekly timed 5 km runs/walks in parks worldwide~\cite{hindley2020}.  A
typical event will have 200 participants.  Table~\ref{parkruntable}
shows the author has completed a total of 15 parkruns to date, and
from the first pair of numbers we see that 238 runners attended that
particular parkrun, of whom the author placed 173.

\begin{table*}[t]
  \caption{Parkrun results}
\label{parkruntable}
\begin{tabular}{ccccccccccc}
\hline
rank   &173  & 165 & 172 & 199 & 181 & 229 & 177 & 222 & 206 & 142 \\
runners&238  & 238 & 196 & 242 & 242 & 318 & 259 & 305 & 297 & 241 \\ \\ \\
rank   & 118 & 224 & 128 & 115 & 183 &     &     &     &     &\\
runners& 179 & 338 & 203 & 245 & 254 &     &     &     &     &\\
\hline
\end{tabular}
\end{table*}

We may consider the author to have an unknown generalized Bradley
Terry strength $a$: Figure~\ref{parkrunsupport} shows a support curve
for $a$ with the evaluate, $\hat{a}\simeq 0.448$ shown.  We also see
two units of support~\cite{edwards1992} shown as a dotted line
illustrating a support interval of about 0.315 - 0.567; we may be
reasonably confident that the author's true strength lies in this
range.  In particular, note that a reasonable $H_0\colon
a=\frac{1}{2}$ may not be rejected, the support for $H_0$ being only
susceptible to very minor improvement [$\simeq 0.351$] by moving to
the evaluate.

\begin{figure}[t]
\includegraphics[width=4in]{plotparkrun}  % plotparkrun.pdf made in very_simplified_likelihood.Rmd
\caption{Author's strength\label{parkrunsupport}}
\end{figure}

\subsection{Education}

One issue encountered in educational research is that of
confidentiality of examination scores, and individual students'
results are generally regarded as sensitive and private information.
Nonetheless, it is common for students to possess two pertinent and
informative observations: the total enrollment in a specific class and
their individual ranking within that class; table~\ref{educationtable}
shows an apocryphal dataset.  What could a student infer from such a
dataset?

\begin{table*}[t]
  \caption{Educational dataset}
\label{educationtable}
\begin{tabular}{llll}
\hline
                   course&category   &rank&class size\\ \hline
      rings and modules  &  pure     & 9  &      12\\
           group theory  &  pure     & 7  &      17\\
               calculus  & applied   & 2  &      23\\
         linear algebra  & applied   & 3  &       9\\
 differential equations  & applied   & 2  &      13\\
               topology  &   pure    & 8  &      14\\
     special relativity  & applied   & 5  &      13\\
        fluid mechanics  & applied   & 4  &      12\\
            Lie algebra  &   pure    & 9  &      15\\
     \hline
\end{tabular}
\end{table*}

One natural question would be to wonder whether the student was bettar
at pure or applied mathematics and, if so, to quantify the difference.
Figure~\ref{plotpureandapplied} shows contours of support [again
  normalised so the support of the evaluate is zero] for the dataset,
as a function of pure strength on the horizontal axis and applied
strength on the vertical axis.  A null of equal strengths appears as a
diagonal line.  We see that the maximum likelihood estimate is
$\hat{p}_\mathrm{pure}=0.567$, $\hat{p}_\mathrm{applied}=0.887$.
However, a constrained optimization shows that the maximum support
along the null diagonal is about $-2.43$.  Edwards's two units of
support criterion is met, and we are justified in concluding that the
student is indeed more competent at applied than pure mathematics.
Alternatively, we may observe that $2\mathcal{S}=4.86$ is in the tail
region of its asymptotic $\chi^2_1$ distribution, corresponding to a
$p$-value of about $0.028$.

\begin{figure}
\includegraphics[width=4in]{plotpureandapplied}  % plotpureandapplied.pdf produced by very_simplified_likelihood.Rmd
\caption{Contours of equal support for pure (horizontal) and applied
  (vertical) strength; null of equal strengths shown as a diagonal
  line  \label{plotpureandapplied}}
\end{figure}

\subsection{Formula 1}

Formula 1 motor racing is an important and prestigious motor sport
\citep{codling2017,jenkins2010}.  Season ranking is based on a points
allocation system wherein competitors are awarded points based on race
finishing order; points accumulate additively.  The overall
competition winner is the competitor who accumulates the most points
after the final race.  However, as argued by
Hankin~\cite{hankin2023_formula1points}, the drivers' {\em ranks} are
statistically sufficient for analysis using Plackett-Luce approaches.
The methods presented here may be used to assess competitive strengths
of Formula 1 racing driver Sergio P\'{e}rez (``Checo'') over his
career from 2011 to 2023.  Checo's Formula 1 performance began
somewhat inauspiciously as second driver for Sauber for whom he
finished no higher than P7 in 2011, but his performance rose
subsequently, culminating in his achievement of a startling 20 podium
finishes in 2022-3.  Noting that Checo raced against 73 distinct
competitors (this rendering estimation of the other players'
Plackett-Luce strengths effectively impossible~\cite{hankin2020}), we
are nevertheless in a position to consider competitive strength using
the modified likelihood approach presented here.  We consider the
following inference problem:

\begin{equation}
  a = a(t) = \frac{e^{\alpha + \beta t}}{1+e^{\alpha + \beta t}},\qquad\alpha,\beta\in\mathbb{R}
\end{equation}

(that is, a logistic regression of generalized Bradley-Terry strength
$a$ against time, with $\alpha$ and $\beta$ being linear
coefficients).  Contours of equal support are shown in
Figure~\ref{showchecolike}, again normalized so that
$\mathcal{S}(\hat{\alpha},\hat{\beta})=0$.  We see the expected
elliptical contours, and further $H_0\colon\beta=0$ may be rejected in
favour of $\alpha=-0.27$, $\beta=0.081$, on the grounds that
$\mathcal{S}(\alpha,0)\leqslant 9.2$.  Alternatively we may observe
that $2\mathcal{S}=18.4$ would correspond to a $p$-value of about
$1.7\times 10^{-5}$ on its asymptotic $\chi^2_1$ null distribution.

\begin{figure}[t]
\includegraphics[width=4in]{showchecolike}  % showchecolike.pdf produced by very_simplified_likelihood.Rmd
\caption{Likelihood functions for $\alpha,\beta$ for Checo's racing
  career 2011-2023\label{showchecolike}}
\end{figure}

\section{Conclusions and further work}

Here a simplification of the Plackett-Luce likelihood functions for
order statistics was proposed, some simple properties obtained, and
inference problems from three disciplines given.  The results seem to
be difficult to obtain any other way.  The assumptions furnish a
reasonable, intuitive, and computationally tractable likelihood
function with wide applicability.  Further work might include the
analysis of more than one focal competitor.

% \begin{acks}[Acknowledgments]
% The authors would like to thank the anonymous referees, an Associate
% Editor and the Editor for their constructive comments that improved the
% quality of this paper.
% \end{acks}

### 2024-03-01 inst/very_simplified_likelihood.Rmd
start to write up for ejs

```{r showfilledc}
L <- pmax(L,-40)
contour(a,b,L)
```

@@

```{r showpvalue}
pchisq(2*(o_free$value - o_constrained$value),df=1,lower.tail=FALSE)
```

@@

```{r definesupportparkrun}
L <- function(a,results){
   out <- a*0
   for(i in seq_len(nrow(results))){
   jj <- f_vec_a(a,r=results$place[i]-1,results$runners[i]-1,log=TRUE)
   out <- out + jj
   }
   out[is.infinite(out)] <- NA
   return(out)
}
```

```{r calculatelikes,cache=F}
a <- seq(from=0.25,to=0.65,by=0.01)
La <- f_vec_arn(a,r=results$place-1,results$runners-1,log=TRUE)
La <- La - max(La,na.rm=TRUE)
```

```{r optimsinglelike}
(optparkrun <- optimize(function(a){L(a,results)},c(0.2,0.8),maximum=TRUE))
f <- function(a){L(a,results)-2}
```

```{r plotparkrunlike, fig.cap="Support curve for author's generalized Bradley-Terry strength.  Evaluate of $\\hat{a}\\simeq 0.448$ has a support of zero"}
plot(a,La,xlab="generalized Bradley-Terry strength",ylab="support")
abline(h=c(0,-2))
abline(v=0.5)
```

Figure \@ref(fig:plotparkrunlike) shows a support curve for the
author's generalized Bradley-Terry strength, normalized so that
$\mathcal{S}(\hat{a})=0$.  We see a credible interval [that is, values
of $a$ for which $\mathcal{S}(a)\geqslant -2$] of about

@@

```{r,label=contourstrength}
par(pty='s')
contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5),
xlab='pure strength',ylab='applied strength')
abline(0,1)
```

>>>>

```{r showfilledc}
L <- pmax(L,-40)
showchec <- function(...){
contour(a,b,L,xlab=expression(alpha),ylab=expression(beta))
}
showchec()
pdf(file="showchecolike.pdf")
contour(a,b,L)
dev.off()
```

@@

```{r showpvalue}
o_free
o_constrained
pchisq(2*(o_free$value - o_constrained$value),df=1,lower.tail=FALSE)
```

@@

```{r optimsinglelike,cache=TRUE}
```

```{r calculate_parksupp,cache=TRUE}
S_parkrun <- function(a){f_vec_arn(a,r=results$place-1,results$runners-1,log=TRUE)}
(optparkrun <- optimize(S_parkrun, c(0.2,0.8),maximum=TRUE))

```{r calcparka,cache=TRUE}
a <- seq(from=0.25,to=0.65,by=0.01)
S_parkrun_a <- S_parkrun(a) - optparkrun$objective
uniroot(function(a){S_parkrun(a) - optparkrun$objective+2},c(0.1,0.5))$root
uniroot(function(a){S_parkrun(a) - optparkrun$objective+2},c(0.5,0.9))$root
optparkrun$objective - S_parkrun(0.5)
```

```{r plotparkrunlike, fig.cap="Support curve for author's generalized Bradley-Terry strength.  Curve is normalised so that the evaluate of $\\hat{a}\\simeq 0.448$ has a support of zero.  Dotted line shows $\\mathcal{S}=-2$; intersection of this with the  curve shows that the support interval runs from about 0.315 to 0.567"}
plotparkrun <- function(...){
plot(a,S_parkrun_a,xlab="generalized Bradley-Terry strength",ylab="support",type='b',ylim=c(-5,0))
abline(h=c(0,-2),lty=2)
segments(x0=0.5,y0=-1.5,y1=0)
segments(x0=optparkrun$maximum,y0=-1,y1=0)
text(optparkrun$maximum,-0.8,expression(hat(a)==0.448),pos=2)
text(0.5,-1.5,expression(H[o]:a==0.5),pos=2)
}
plotparkrun()
pdf("plotparkrun.pdf")
plotparkrun()
dev.off()
```

Figure \@ref(fig:plotparkrunlike) shows a support curve for the
author's generalized Bradley-Terry strength, normalized so that
$\mathcal{S}(\hat{a})=0$.  We see a support interval [that is,
$\left\lbrace a\colon\mathcal{S}(a)\geqslant -2\right\rbrace$] of
about 0.315 to 0.567.  We might also observe that the support for
$H_0\colon a=\frac{1}{2}$ [that is, that the author has a 50\% chance
of running faster than a randomly chosen Parkrun competitor] is about
0.35, less than Edwards's two units of support criterion.  We thus see
some support for Hankin's assertion that he is "as fit as anyone
else there".

@@

```{r,label=contourstrength}
plotpureandapplied <- function(...){
par(pty='s')
contour(x,x,L,asp=1,xlim=range(x),ylim=range(x),levels=-(0:5),
xlab='pure strength',ylab='applied strength')
abline(0,1)
}
plotpureandapplied()
pdf(file="plotpureandapplied.pdf")
plotpureandapplied()
dev.off()
```
